---
title: "Hierarchical Models"
author: "Haohan Chen"
date: "March 22, 2018"
output: html_document
---

```{r, message=FALSE}
# !! Please run this chunk before you compile the document !!

#-----------------------------
# install required packages
#-----------------------------
pkgs <- c("lme4", "brms", "arm", "ggplot2")

usePackage <- function(p) 
{
  for (pkg in p){
    if (!is.element(pkg, installed.packages()[,1]))
      install.packages(pkg, dep = TRUE)
    require(pkg, character.only = TRUE)
  }
}

usePackage(pkgs)
```

# Example: Who Support Trump in 2016 Election?

Data: ANES 2016, version 2017. 

```{r, eval=FALSE, include=FALSE}
# Some data preprocessing using the raw data. 
# I've uploaded the cleaned dataset. This is just FYI.

# Data Cleaning
d <- read.table("anes_timeseries_2016/anes_timeseries_2016_rawdata.txt", header = T, sep = "|")

var <- list()

# Extract variables

# -------------
# Y
# -------------
# Case ID
  var$V160001 <- d$V160001
  
# Turnout: 1. yes; 2. no. -6 no interview
  var$turnout <- as.numeric(d$V162034 == 1)
  
# Vote: 1. Clinton; 2. Trump; -6 no interview
  var$vote_clinton <- as.numeric(d$V162034a == 1)
  var$vote_trump <- as.numeric(d$V162034a == 2)
  
# Thermometer: -99 refused
  var$therm_dem_cand <- d$V162078
  var$therm_dem_cand[var$therm_dem_cand < 0] <- NA
  var$therm_rep_cand <- d$V162079
  var$therm_rep_cand[var$therm_rep_cand < 0] <- NA
# -------------
# X
# -------------
# Demographics
  # Age: -9 refused
  var$age <- d$V161267
  var$age[var$age == -9] <- NA
  
  # Marital status
  var$married <- as.numeric(d$V161268 == 1)
  
  # Children
  var$n_children <- d$V161324
  var$n_children[var$n_children == -9] <- NA
  
  # social class
    # 1. Lower class 64 213 277
    # 2. Working class 191 681 872
    # 3. Middle class 298 1071 1369
    # 4. Upper class 24 95 119
    # -1. Inapplicable 596 1013 1609
    # -8. Don’t know 4 0 4
    # -9. Refused 4 17 21
  var$class <- d$V161307
  
  # Race
    # 1. White, non-Hispanic 796 2242 3038
    # 2. Black, non-Hispanic 120 278 398
    # 3. Asian, native Hawaiian or
    # other Pacif Islr,non-Hispanic 30 118 148
    # 4. Native American or Alaska
    # Native, non-Hispanic 10 17 27
    # 5. Hispanic 169 281 450
    # 6. Other non-Hispanic incl
    # multiple races <WEB: blank
    # ’Other’ counted as a race> 52 125 177
    # -9. Refused 4 29 33
  var$race <- factor(d$V161310x)

  # Education
    # 13. Bachelor’s degree (for
    # example: BA, AB, BS) 265 690 955
    # 10. Some college but no
    # degree 205 694 899
    # 9. High school graduate- high
    # school diploma or equivalent
    # (for example: GED) 258 552 810
    # 14. Master’s degree (for
    # example: MA, MS, MENG,
    # MED, MSW, MBA) 120 379 499
    # 11. Associate degree in
    # college -
    # occupational/vocational
    # program 101 212 313
    # 12. Associate degree in
    # college – academic program 83 205 288
    # 8. 12th grade no diploma 40 67 107
    # 16. Doctorate degree (for
    # example: PHD, EDD) 19 74 93
    # 15. Professional school degree
    # (for example: MD, DDS,
    # DVM, LLB, JD) 17 71 88
    # 7. 11th grade 24 38 62
    # 6. 10th grade 13 27 40
    # 5. 9th grade 8 24 32
    # 95. Other SPECIFY 7 17 24
    # 4. 7th or 8th grade 8 14 22
    # -9. Refused 0 15 15
  var$edu <- d$V161270
  var$edu[var$edu == -9] <- NA
# State: address  
  var$state <- factor(d$V161010e)

d_anes <- as.data.frame(var)
d_anes <- d_anes[complete.cases(d_anes),]
save(d_anes, file = "d_anes.Rdata")
rm(list=ls())
```

```{r}
# Load data 
load("d_anes.Rdata")
```

# Varying Intercept, Same Slope Across Pools

## Continuous Response: Feeling Thermometer about Republican Candidate

### Pooling
```{r}
m.pooled <- lm(therm_rep_cand ~ age + married + edu, data = d_anes)
summary(m.pooled)
```

### Un-Pooled
```{r}
class(d_anes$state)
m.unpooled <- lm(therm_rep_cand ~ age + married + edu + state - 1, data = d_anes)
# Simulate the Fixed Effect (same trick as what we've done)
summary(m.unpooled)
length(coef(m.unpooled))
mu <- coef(m.unpooled)[4:54]
v <- vcov(m.unpooled)[4:54, 4:54]
fe_sim <- MASS::mvrnorm(10000, mu = mu, Sigma = v)
fe_sim_95CI <- as.data.frame(t(apply(fe_sim, 2, function(x) quantile(x, c(0.025, 0.5, 0.975)))))
fe_sim_95CI$state <- substr(row.names(fe_sim_95CI), 6, 7)
# Plot Fixed Effect
library(ggplot2)
ggplot(fe_sim_95CI, aes(y = `50%`, x = state)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Feeling Thermometer (Trump)") + xlab("State") + coord_flip() 

# Can't compare among states? Sort the level of state: from highest to lowest
fe_sim_95CI$state_ordered <- factor(fe_sim_95CI$state, levels = fe_sim_95CI$state[order(fe_sim_95CI$`50%`)])

ggplot(fe_sim_95CI, aes(y = `50%`, x = state_ordered)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Feeling Thermometer (Trump)") + xlab("State") + coord_flip() 

# Anything wierd in this graph?
  # Check the variance!
  # Check the number of respondent
sort(table(d_anes$state))
fe_sim_95CI$`Few Cases` <- fe_sim_95CI$state %in% names(sort(table(d_anes$state))[1:5])

ggplot(fe_sim_95CI, aes(y = `50%`, x = state_ordered, color = `Few Cases`)) + geom_point(aes(shape = `Few Cases`, size = `Few Cases`)) + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Feeling Thermometer (Trump)") + xlab("State") + coord_flip() 
fe_sim_95CI$`Few Cases` <- NULL
# Solution: Hierarchical Model
```

### Partial Pooling

#### Frequentist Hierarchical Model
```{r}
# Fit a Hierarchical model
m.part_pool <- lmer(therm_rep_cand ~ age + married + edu - 1 + (1|state), data = d_anes)
summary(m.part_pool)

# Note: $state select the random effect of interest. you have multiple sometimes
re_sim_95CI <- data.frame(
  `2.5%` = ranef(m.part_pool)$state + qnorm(0.025) * se.ranef(m.part_pool)$state,
  `50%` = ranef(m.part_pool)$state,
  `97.5%` = ranef(m.part_pool)$state + qnorm(0.975) * se.ranef(m.part_pool)$state,
  state = rownames(ranef(m.part_pool)$state)
)
names(re_sim_95CI) <- c("2.5%", "50%", "97.5%", "state")

# Plot
ggplot(re_sim_95CI, aes(y = `50%`, x = state)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Feeling Thermometer (Trump)") + xlab("State") + coord_flip() 

# Can't compare among states? Sort the level of state: from lowest to highest
re_sim_95CI$state_ordered <- factor(re_sim_95CI$state, levels = re_sim_95CI$state[order(re_sim_95CI$`50%`)])
ggplot(re_sim_95CI, aes(y = `50%`, x = state_ordered)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Feeling Thermometer (Trump)") + xlab("State") + coord_flip() 

# Compare the two
combine_re_fe <- rbind(fe_sim_95CI, re_sim_95CI)
combine_re_fe$type <- rep(c("Unpooled", "Partial-Pooled"), each = 51)
# Compare Unpooled and Partial-pooled results
ggplot(combine_re_fe, aes(y = `50%`, x = state, color = type)) + 
  geom_point(aes(shape = type), position = position_dodge(width = 0.5)) + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`), position = position_dodge(width = 0.5), width = 0.3) + 
  ylab("Feeling Thermometer (Trump)") + xlab("State") + coord_flip()

# Compare Unpooled and Partial-pooled results
ggplot(combine_re_fe, aes(y = `50%`, x = state_ordered, color = type)) + 
  geom_point(aes(shape = type), position = position_dodge(width = 0.5)) + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`), position = position_dodge(width = 0.5), width = 0.3) + 
  ylab("Feeling Thermometer (Trump)") + xlab("State") + coord_flip()

```

#### Bayesian Hierarchical Model
```{r}
# BRMS package
# https://cran.r-project.org/web/packages/brms/vignettes/brms_overview.pdf

m.part_pool_bayes <- brm(therm_rep_cand ~ age + married + edu - 1 + (1|state), data = d_anes,
                         prior = c(set_prior("normal(0, 2)", class = "b")), family = gaussian(identity),
                         warmup = 500, iter = 2000, chains = 2,
                         control = list(adapt_delta = 0.99))

bayes_samp_coef = posterior_samples(m.part_pool_bayes)
colnames(bayes_samp_coef)
bayes_sim_95CI <- as.data.frame(t(apply(bayes_samp_coef[, 6:56], 2, function(x) quantile(x, c(0.025, 0.5, 0.975)))))
bayes_sim_95CI$state <- substr(row.names(bayes_sim_95CI), 9, 10)

# Plot
ggplot(bayes_sim_95CI, aes(y = `50%`, x = state)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Feeling Thermometer (Trump)") + xlab("State") + coord_flip() 

# Can't compare among states? Sort the level of state: from lowest to highest
bayes_sim_95CI$state_ordered <- factor(bayes_sim_95CI$state, levels = bayes_sim_95CI$state[order(bayes_sim_95CI$`50%`, decreasing = T)])
ggplot(bayes_sim_95CI, aes(y = `50%`, x = state_ordered)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Feeling Thermometer (Trump)") + xlab("State") + coord_flip() 


# Compare 
combine_re_fe_bayes <- rbind(combine_re_fe, cbind(bayes_sim_95CI, type = "Bayes"))

ggplot(combine_re_fe_bayes, aes(y = `50%`, x = state_ordered, color = type)) + 
  geom_point(aes(shape = type), position = position_dodge(width = 0.5)) + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`), position = position_dodge(width = 0.5), width = 0.3) + 
  ylab("Feeling Thermometer (Trump)") + xlab("State") + coord_flip()
```



## Categorical Outcome: Voted for Trump

```{r}
# Load data 
rm(list=ls())
load("d_anes.Rdata")
```


### Pooling
```{r}
m.pooled <- glm(vote_trump ~ age + married + edu, family = binomial(link = "logit"), data = d_anes)
summary(m.pooled)
```

### Un-Pooled
```{r}
m.unpooled <- glm(vote_trump ~ age + married + edu + state - 1, family = binomial(link = "logit"), data = d_anes)
# Simulate the Fixed Effect (same trick as what we've done)
length(coef(m.unpooled))
mu <- coef(m.unpooled)[4:54]
v <- vcov(m.unpooled)[4:54, 4:54]
fe_sim <- MASS::mvrnorm(10000, mu = mu, Sigma = v)
fe_sim_95CI <- as.data.frame(t(apply(fe_sim, 2, function(x) quantile(x, c(0.025, 0.5, 0.975)))))
fe_sim_95CI$state <- substr(row.names(fe_sim_95CI), 6, 7)
# Plot Fixed Effect
library(ggplot2)
ggplot(fe_sim_95CI, aes(y = `50%`, x = state)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Coefficient") + xlab("State") + coord_flip() 

# Can't compare among states? Sort the level of state: from highest to lowest
fe_sim_95CI$state_ordered <- factor(fe_sim_95CI$state, levels = fe_sim_95CI$state[order(fe_sim_95CI$`50%`)])
ggplot(fe_sim_95CI, aes(y = `50%`, x = state_ordered)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Coefficient") + xlab("State") + coord_flip() 

# Anything wierd in this graph?
  # Check the variance!
  # Check the number of respondent
sort(table(d_anes$state))
fe_sim_95CI$`Few Cases` <- fe_sim_95CI$state %in% names(sort(table(d_anes$state))[1:5])

ggplot(fe_sim_95CI, aes(y = `50%`, x = state_ordered, color = `Few Cases`)) + geom_point(aes(shape = `Few Cases`, size = `Few Cases`)) + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Coefficient") + xlab("State") + coord_flip() 
fe_sim_95CI$`Few Cases` <- NULL
# Solution: Hierarchical Model
```

### Partial-Pooling

#### Frequentist
```{r}
# Fit a Hierarchical model
m.part_pool <- glmer(vote_trump ~ age + married + edu - 1 + (1|state), family = binomial(link = "logit"), data = d_anes)
summary(m.part_pool)

# Note: $state select the random effect of interest. you have multiple sometimes
re_sim_95CI <- data.frame(
  `2.5%` = ranef(m.part_pool)$state + qnorm(0.025) * se.ranef(m.part_pool)$state,
  `50%` = ranef(m.part_pool)$state,
  `97.5%` = ranef(m.part_pool)$state + qnorm(0.975) * se.ranef(m.part_pool)$state,
  state = rownames(ranef(m.part_pool)$state)
)
names(re_sim_95CI) <- c("2.5%", "50%", "97.5%", "state")

# Plot
ggplot(re_sim_95CI, aes(y = `50%`, x = state)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Coefficient") + xlab("State") + coord_flip() 

# Can't compare among states? Sort the level of state: from lowest to highest
re_sim_95CI$state_ordered <- factor(re_sim_95CI$state, levels = re_sim_95CI$state[order(re_sim_95CI$`50%`, decreasing = T)])
ggplot(re_sim_95CI, aes(y = `50%`, x = state_ordered)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Coefficient") + xlab("State") + coord_flip() 

# Compare the two
combine_re_fe <- rbind(fe_sim_95CI, re_sim_95CI)
combine_re_fe$type <- rep(c("Unpooled", "Partial-Pooled"), each = 51)
# Compare Unpooled and Partial-pooled results
ggplot(combine_re_fe, aes(y = `50%`, x = state, color = type)) + 
  geom_point(aes(shape = type), position = position_dodge(width = 0.5)) + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`), position = position_dodge(width = 0.5), width = 0.3) + 
  ylab("Coefficient") + xlab("State") + coord_flip()

# Compare Unpooled and Partial-pooled results
ggplot(combine_re_fe, aes(y = `50%`, x = state_ordered, color = type)) + 
  geom_point(aes(shape = type), position = position_dodge(width = 0.5)) + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`), position = position_dodge(width = 0.5), width = 0.3) + 
  ylab("Coefficient") + xlab("State") + coord_flip()
```

#### Bayesian
```{r}
m.part_pool_bayes <- brm(vote_trump ~ age + married + edu - 1 + (1|state), data = d_anes, 
                         prior = c(set_prior("normal(0, 2)", class = "b")), family = bernoulli(link = "logit"),
                         warmup = 500, iter = 2000, chains = 2,
                         control = list(adapt_delta = 0.99))

bayes_samp_coef = posterior_samples(m.part_pool_bayes)
colnames(bayes_samp_coef)
# Subset state intervals. Then get 95% CI.
bayes_sim_95CI <- as.data.frame(t(apply(bayes_samp_coef[, 6:55], 2, function(x) quantile(x, c(0.025, 0.5, 0.975)))))
bayes_sim_95CI$state <- substr(row.names(bayes_sim_95CI), 9, 10)

# Plot
ggplot(bayes_sim_95CI, aes(y = `50%`, x = state)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Coefficient") + xlab("State") + coord_flip() 

# Can't compare among states? Sort the level of state: from lowest to highest
bayes_sim_95CI$state_ordered <- factor(bayes_sim_95CI$state, levels = bayes_sim_95CI$state[order(bayes_sim_95CI$`50%`, decreasing = T)])
ggplot(bayes_sim_95CI, aes(y = `50%`, x = state_ordered)) + geom_point() + 
  geom_errorbar(aes(ymin = `2.5%`, ymax = `97.5%`)) +
  ylab("Coefficient") + xlab("State") + coord_flip()
```