---
title: "`Rmarkdown` for Data Analysis: A Refresher"
author: "Haohan Chen^[Political Science Department, Duke University. haohan.chen@duke.edu]"
date: "February 2, 2018"
fontsize: 11pt
geometry: margin=1cm
linkcolor: blue
output: 
  pdf_document:
    number_sections: false
    keep_tex: true
    fig_caption: true
    citation_package: natbib
header-includes:
#  - \usepackage{TikZ}
#  - \usepackage{times}
---


This is a refresher of some common `Rmarkdown` operations, to help you editing your homework and papers more efficiently for PS733. I demonstrate it by writing a mini data analysis report on a toy dataset.

# `Rmarkdown` Setup

```{r setup}
  # enable setting font size of code chunk
  def.chunk.hook  <- knitr::knit_hooks$get("chunk")
  knitr::knit_hooks$set(chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(options$size != "normalsize", 
           paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
  })
  
  # knitr options
  knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                        results = "hold",
                        fig.path = "figures/", size = "small")
  # Explanation in the following chunk -- when fontsize is reduced!
```

```{r setup2}
#-----------------------------
# About the Header
#-----------------------------
  # keep_tex: output the tex file (so that you can directly use the tex code generated)
  # fig_caption: show caption of figures. true by default.
  # citation_package: use latex natbib citation package for bibliography. recommended!
  # header_includes: include other command the document's preamble.
  #                  mostly used it to call more LaTeX packages.
#-----------------------------
# Global options
#-----------------------------
  # About the font size mess
    # A trick that enables you to custimize the fontsize of code in the chunk.
    # I have to do this because Rmarkdown does not directly support setting

  # About knitr::opt_chunk
    # echo: show code
    # message, warning: show system generated info (e.g. progress bar)
    # results = "hold". hold output of results till the end of chunk (invalid for fig)
    # fig.path: set a path to store figures generate. can reuse them elsewhere.
    #           without this, no fig will be saved.
    # size: font size of *code in the chunks* (not your main text, which is set
    #       in the header "fontsize: 11pt". options of size include "small", "tiny",
    #       "normalisize", "huge"...

  # Also, Create a directory to save your tables (used later)
  dir.create("tables")
  # Output type of this file is LaTex (a param for later)
  out_type = "latex"
```

# Packages and Dataset Setup

```{r package}
#-----------------------------
# load/install required packages
#-----------------------------
# Names of all packages used
  pkgs <- c("dplyr", "ggplot2", "xtable", "stargazer", "PerformanceAnalytics", "cowplot")
# A function to load all above packages. Install if they have not been installed.
  usePackage <- function(p){
    for (pkg in p){
      if (!is.element(pkg, installed.packages()[,1]))
        install.packages(pkg, dep = TRUE, repos = "https://cloud.r-project.org/")
      require(pkg, character.only = TRUE)
    }
  }
  usePackage(pkgs)
```

```{r loaddata}
#-----------------------------
# load your data
#-----------------------------
  # Load your dataset of interest. 
  # Below is an example economic dataset coming with R
    data("longley")
      # J. W. Longley (1967) An appraisal of least-squares programs from 
      # the point of view of the user. 
      # Journal of the American Statistical Association 62, 819-841.
  # Just to mess up the dataset by a bit
    names(longley) <- c("gnp.def", "gnp", "unemp", "force", "pop", "yr", "emp")
```

\clearpage

# Exploratory Data Analysis

Table \ref{tab:desc} shows the descriptive statistics. Figure \ref{fig:corrmat} is the Correlation Matrix. Figure \ref{fig:force-and-gnp1} shows the relationship between GNP and the size of armed force using the default `plot` function. Figure \ref{fig:force-and-gnp2} is the same plot using `ggplot`. 

## Table

```{r edatab1}
#---------------------------
# Table of summary statistics
#---------------------------
  # Summary statistics
    summary(longley)  
    # Not pretty. We can do better!
```

```{r edatab2, results='asis'}
#-------------------------------------
# Table of summary statistics (con'd)
#-------------------------------------
  # Produce a LaTeX summary stats table (can also be HTML)
  stargazer(longley, title = "Descriptive Statistics",
      mean.sd = TRUE, median = TRUE, iqr = TRUE, min.max = TRUE,
      header = FALSE, label = "tab:desc", type = out_type)
    # Will come back to Stargazer soon.
```

### Tip: Save your Table
I recommend saving your table in a separate `.tex` file for convenient re-use.
```{r edatab3, results='hide'}
  # Instead of directly output your outcome. 
  # Saving the output is a better strategy. Think about why.
  desc_tab <- capture.output(
    stargazer(longley, title = "Descriptive Statistics",
        mean.sd = TRUE, median = TRUE, iqr = TRUE, min.max = TRUE,
        header = FALSE, label = "tab:desc", type = out_type)
    )
  # Save it to a folder for tables (created earlier)
  writeLines(desc_tab, "tables/descriptive.tex")    
```

\input{tables/descriptive}

<!-- Note: text within these two symbols are "comment", not included in either the text or the code. -->
<!-- Note: if you use "\include", it will put the table in a new page --> 

\clearpage

## Correlcation Matrix
```{r cormat, fig.width=8, fig.height=6, fig.align='center', out.width="100%", fig.pos="h!", fig.cap="\\label{fig:corrmat}Correlation Matrix"}
#---------------------------
# Correlation Matrix
#---------------------------
  PerformanceAnalytics::chart.Correlation(longley)
  # By far my favorite, better than other fancy stuff.
  # Perfect for continuous variables
```

## Correlation Plots (and their arrangement)

```{r cor1, fig.cap="\\label{fig:force-and-gnp1}Size of Armed Force and GNP (default)", fig.width=8, fig.height=4, fig.align='center', out.width="90%", fig.pos="h!"}
par(mfrow = c(1, 2)) # 2 figures in a row
plot(longley$gnp, longley$force, xlab = "GNP", ylab = "Size of Armed Force", main = "GNP")
plot(log(longley$gnp), longley$force, xlab = "log(GNP)", ylab = "Size of Armed Force", 
     main = "log(GNP)")
```

```{r cor2, fig.width=8, fig.height=4, fig.align='center', out.width="90%", fig.pos="h!", fig.cap="\\label{fig:force-and-gnp2}Size of Armed Force and GNP (ggplot)"}
# The cowplot package: https://cran.r-project.org/web/packages/cowplot/vignettes/plot_grid.html
fig_cor1 <- ggplot(longley, aes(x = gnp, y = force)) + geom_point() + 
  geom_smooth(method = "loess") + xlab("GNP") + ylab("Size of Armed Force") +
  ggtitle("GNP")
fig_cor2 <- ggplot(longley, aes(x = log(gnp), y = force)) + geom_point() + 
  geom_smooth(method = "loess") + xlab("GNP") + ylab("Size of Armed Force") +
  ggtitle("log(GNP)")
plot_grid(fig_cor1, fig_cor2, ncol = 2)
```


\clearpage

# Models

Clearly state your model and the assumption of the model.

(Alignment Style 1:)
\begin{gather*}
  \text{Model 1: } \quad \text{Armed Force}_i = \beta_0 + \beta_1 \text{Unemployment}_i + \beta_2 \text{GNP}_i + \epsilon_i \\
  \text{Model 2: } \quad \text{Armed Force}_i = \beta_0 + \beta_1 \text{Unemployment}_i + \beta_2 \text{GNP}_i + \beta_3 \text{Population}_i + \epsilon_i \\
  \text{Model 3: } \quad \text{Armed Force}_i = \beta_0 + \beta_1 \text{Unemployment}_i + \beta_2 \text{GNP}_i  + \beta_3 \text{GNP}_i^2 + \beta_4 \text{Population}_i + \epsilon_i \\
  \text{Model 4: } \quad \text{Armed Force}_i = \beta_0 + \beta_1 \text{Unemployment}_i + \beta_2 \text{GNP}_i  + \beta_3 \text{GNP}_i^2 + \beta_4 \text{Population}_i + \beta_5 \text{Year}_i + \epsilon_i \\
  \text{For all models, I assume }  \epsilon \sim N(0, \sigma^2)
\end{gather*}

(Alignment Style 2:)
\begin{align*}
  \text{Model 1: } & \text{Armed Force}_i = \beta_0 + \beta_1 \text{Unemployment}_i + \beta_2 \text{GNP}_i + \epsilon_i \\
  \text{Model 2: } & \text{Armed Force}_i = \beta_0 + \beta_1 \text{Unemployment}_i + \beta_2 \text{GNP}_i + \beta_3 \text{Population}_i + \epsilon_i \\
  \text{Model 3: } & \text{Armed Force}_i = \beta_0 + \beta_1 \text{Unemployment}_i + \beta_2 \text{GNP}_i  + \beta_3 \text{GNP}_i^2 + \beta_4 \text{Population}_i + \epsilon_i \\
  \text{Model 4: } & \text{Armed Force}_i = \beta_0 + \beta_1 \text{Unemployment}_i + \beta_2 \text{GNP}_i  + \beta_3 \text{GNP}_i^2 + \beta_4 \text{Population}_i + \beta_5 \text{Year}_i + \epsilon_i
\end{align*}
\begin{equation*}
  \text{For all models, I assume } \epsilon \sim N(0, \sigma^2)
\end{equation*}

```{r fit}
#---------------------------
# Fit models
#---------------------------
  # Tips: store a group of model in a list
  # Benefits: convenient management! 
  fit_models <- function(d){
    m <- list()
    m[["Baseline"]] <- glm(force ~ unemp + gnp, data = d, family = gaussian)
    m[["Population"]] <- glm(force ~ unemp + gnp + pop, data = d, family = gaussian)
    m[["Quad Population"]] <- glm(force ~ unemp + gnp + I(gnp^2) + pop, data = d, 
                                  family = gaussian)
    m[["Year"]] <- glm(force ~ unemp + gnp + I(gnp^2) + pop + yr, data = d, 
                       family = gaussian)
    m
  }

  m <- fit_models(longley)
```

\clearpage

# Results (Tables)

Table \ref{tab:arm1} reports all models with no labels. Table \ref{tab:arm2} reports part of the models. Table \ref{tab:arm3} label the variables, reset the number of digits to report etc.

```{r table, results='asis'}
# Stargazer Quick Reference: https://www.jakeruss.com/cheatsheets/stargazer/
# Alternative: xtable. More flexible, but harder to code.
    # https://cran.r-project.org/web/packages/xtable/vignettes/xtableGallery.pdf

#-------------------------------------
# Show regression results with tables
#-------------------------------------
  # Print all models
    stargazer(m, label = "tab:arm1", 
              title = 
                "(All Models) Economic Determinants of the Size of Armed Force",
              header = FALSE, type = out_type)
      # If you input a list of models, it will report them all in one table.
      # Remember to add label and title to your table. 
      # A table of ambiguous meaning is not worth reporting

  # Print a subset of models
    stargazer(m[["Baseline"]], m[["Population"]], label = "tab:arm2", 
              title = 
                "(Baseline and Population) Economic Determinants of the Size of Armed Force",
              header = FALSE, type = out_type)

  # Label your Table (Essential!!!)
    stargazer(m, label = "tab:arm3",
              title = "(Labeled) Economic Determinants of the Size of Armed Force",
              covariate.labels = c("Unemployment", "GNP",
                                   "GNP sq", "Population", "Year"),
                # Mind the order... Better Strategy is asigning meaningful var names
                # in the dataset. Will end up saving your time!
              dep.var.labels = "Size of Armed Force",
              digits = 2,
              ci = TRUE,
              star.cutoffs = NA, # don't show stars
              notes = "Source of Data: Longley (1967)",
              font.size = "footnotesize", # Font size
              header = FALSE, type = out_type
              )
```

\clearpage

# Regression Results (Graphs)

**This part is muted by default because of the long time it takes to install the package.**

Figure \ref{fig:coef1a} and \ref{fig:coef1b} plots the coefficients with unscaled predictors. Figure \ref{fig:coef2a} and \ref{fig:coef2b} plot the coefficients with  scaled predictors. The interpretation of the coefficients are different after scaling. Before scaling: "one unit change of `variable x` is associated with `coef` units change of the dependent variables". After scaling: "one standard deviation change of `variable x` is associated with `coef` units change of the dependent variables".

In addition, Figure \ref{fig:slope} shows how the dependent variable changes in response to all four independent variables in our `Year` Model.

\textcolor{red}{\textbf{ALL THESE FIGURES ARE TERRIBLE, BECAUSE THEY ARE NOT CORRECTLY LABELED!!}} So I would still recommend using `ggplot` for your visualization, where you enjoy higher flexibility. 

```{r}
# Show the code in the appdx, but do not run them again.
knitr::opts_chunk$set(eval = FALSE)
```

```{r figcoef1, fig.width=6, fig.height=5, fig.align='center', out.width="50%", fig.cap="\\label{fig:coef1a}Coefficients with Unscaled Predictors (one model)", fig.pos="h"}

# Experimenting a good visualization package.
# Takes a long time to install. And there's been error report
# Save your other R files. When you are ready to try, set the above
# eval = TRUE
usePackage("sjPlot")

#---------------------------
# Visualize Regression Results
#---------------------------
  # sjPlot: http://www.strengejacke.de/sjPlot/; 
  #         https://cran.r-project.org/web/packages/sjPlot/vignettes/sjpglm.html
  #         https://github.com/strengejacke/sjPlot
  # Pros: Can work with models generated by different packages
  #       lmer, glm, glmm, glmer, etc.
  #       it's plot appears much better than many default plot functions.
  # Cons: Not flexible compared to ggplot. Package under development. Use with caution.

  #------------------------------
  # Coefficients (original data)
  #------------------------------
    # Plotting one model
    sjPlot::plot_model(m[["Year"]], 
                       title = "Economic Determinants of Armed Force Size I")
```

```{r figcoef2, fig.width=6, fig.height=5, fig.align='center', out.width="50%", fig.cap="\\label{fig:coef1b}Coefficients with Unscaled Predictors (all models)"}
    # Comparing multiple models
    sjPlot::plot_models(m, title = "Economic Determinants of Armed Force Size II")
    # Note. Scale problem --> can't really see CI of most variables.

    # Solution: scale independentvariables before input. BUT DON'T SCALE THE DVs!
```

```{r figcoef3, fig.width=6, fig.height=5, fig.align='center', out.width="50%", fig.cap="\\label{fig:coef2a}Coefficients with Scaled Predictors (one model)"}
  #-----------------------------------------------
  # Coefficients (scaled Independent variables)
  #-----------------------------------------------
    longley_s <- longley %>% mutate(gnp = scale(gnp), unemp = scale(unemp),
                                    pop = scale(pop), yr = scale(yr))
    m <- NULL
    m <- fit_models(longley_s)
    # Plotting one model
    sjPlot::plot_model(m[["Year"]], 
                       title = "Economic Determinants of Armed Force Size I (Scaled)")
```

```{r figcoef4, fig.width=6, fig.height=5, fig.align='center', out.width="50%", fig.cap="\\label{fig:coef2b}Coefficients with Scaled Predictors (all models)"}
    # Comparing multiple models
    sjPlot::plot_models(m, title = "Economic Determinants of Armed Force Size II (Scaled)")
    # Something is wrong with this command. Still need a fix.
```    

```{r figslope, fig.width=8, fig.height=6, fig.align='center', out.width="80%", fig.cap="\\label{fig:slope}How the Dependent Variable Change with All Predictors"} 
  #-----------------------------------------------
  # Other Plots
  #-----------------------------------------------
    sjPlot::plot_model(m[["Year"]], type = "slope")
      # Again. package under developement. 
      # Don't trust it too much. Esp. the Variance.
      # Good for perliminary analysis
```

\clearpage

# Discussion

All results are summarized in Table \ref{tab:arm3}... bla bla bla

# Citation

```{r, echo=FALSE, eval=TRUE}
# I just generate a .bib file (inline citation is a pain) in the same folder as the .Rmd code with this chunk. 
# Substitute it with your real bibliography file.
# Remember. NO SPECIAL CHARACTER in your .bib file. Ideally only a-z,A-Z
bib_text <- '
  @article{stegmueller2013many,
  title={How many countries for multilevel modeling? A comparison of frequentist and Bayesian approaches},
  author={Stegmueller, Daniel},
  journal={American Journal of Political Science},
  volume={57},
  number={3},
  pages={748--761},
  year={2013},
  publisher={Wiley Online Library}
  }

  @article{johnston2014ideology,
  title={Ideology, the Affordable Care Act Ruling, and Supreme Court Legitimacy},
  author={Johnston, Christopher D and Hillygus, D Sunshine and Bartels, Brandon L},
  journal={Public Opinion Quarterly},
  volume={78},
  number={4},
  pages={963--973},
  year={2014},
  publisher={Oxford University Press UK}
  }
  
  @book{beramendi2008democracy,
  title={Democracy, Inequality, and Representation in Comparative Perspective},
  author={Beramendi, Pablo and Anderson, Christopher J},
  year={2008},
  publisher={Russell Sage Foundation}
  }

  @article{bell2015explaining,
  title={Explaining fixed effects: Random effects modeling of time-series cross-sectional and panel data},
  author={Bell, Andrew and Jones, Kelvyn},
  journal={Political Science Research and Methods},
  volume={3},
  number={1},
  pages={133--153},
  year={2015},
  publisher={Cambridge University Press}
  }
'
writeLines(bib_text, "biblio.bib")

```

Bla bla bla \citep{johnston2014ideology}. 

\citet[][p.234]{beramendi2008democracy} argue that... 

Existing studies find evidence that bla bla bla \citep[see][for detailed explanation]{stegmueller2013many,bell2015explaining}...

\bibliography{biblio}

\clearpage

# Others (Analytical Graphs, Game Trees...)

`Rmarkdown` allows you to use all LaTex packages (put `header_includes: \usepackage{} in in the header at the start of the document`). For example, you can plot analytical graphs (functions, game trees etc.) with the `TikZ` packages. See more examples here:\newline
`http://www.sfu.ca/~haiyunc/notes/Game_Trees_with_TikZ.pdf`;\newline `https://sites.google.com/site/kochiuyu/Tikz`.

<!--
## Example: Utility Maximization

\begin{center}
\begin{tikzpicture}[scale=0.6]
  % from https://sites.google.com/site/kochiuyu/Tikz
  \draw[thick,<->] (0,10) node[above]{$y$}--(0,0)--(10,0) node[right]{$x$};
  \node [below left] at (0,0) {$0$};
  \node [below] at (5,0) {$x^{*}$};
  \node [left] at (0,5) {$y^{*}$};
  \draw(1,9)--(9,1) node[right]{$p_xx+p_yy=I$};
  \draw(0,5)--(5,5)--(5,0);
  \draw(2,10) ..controls (3.33,6) and (6,3.33) .. (10,2) node[right]{$U(x,y)=U_0$};
\end{tikzpicture}
\end{center}

## Example: A Simple Sequential Game
\begin{center}
\begin{tikzpicture}[scale=1.5,font=\footnotesize]
  % from http://www.sfu.ca/~haiyunc/notes/Game_Trees_with_TikZ.pdf
  \tikzstyle{solid node}=[circle,draw,inner sep=1.5,fill=black]
  \tikzstyle{hollow node}=[circle,draw,inner sep=1.5]
  \tikzstyle{level 1}=[level distance=15mm,sibling distance=3.5cm]
  \tikzstyle{level 2}=[level distance=15mm,sibling distance=1.5cm]
  \tikzstyle{level 3}=[level distance=15mm,sibling distance=1cm]
  \node(0)[solid node,label=above:{$P1$}]{}
  child{node[solid node,label=above left:{$P2$}]{}
  child{node[hollow node,label=below:{$(1,2)$}]{} edge from parent node[left]{$C$}}
  child{node[hollow node,label=below:{$(1,-1)$}]{} edge from parent node[left]{$D$}}
  child{node[hollow node,label=below:{$(0,2)$}]{} edge from parent node[right]{$E$}}
  edge from parent node[left,xshift=-5]{$A$}
  }
  child{node[solid node,label=above right:{$P2$}]{}
  child{node[hollow node,label=below:{$(2,2)$}]{} edge from parent node[left]{$F$}}
  child{node[hollow node,label=below:{$(1,3)$}]{} edge from parent node[right]{$G$}}
  edge from parent node[right,xshift=5]{$B$}
};
\end{tikzpicture}
\end{center}

\clearpage
-->

# Appendix (Code)

For readability, you may suppress your code within your text, and put them all into the appendix. You can re-use a chunk of code by calling `ref.label=(chunck_name)`. When you reuse a chunk, you may want to avoid running again by setting `eval=FALSE`. Again, you can set these up as a global option with the `knitr::opts_chunck` command.

```{r setupapdx}
# Show the code in the appdx, but do not run them again.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Loading the Data
```{r apdx1, ref.label="loaddata"}
```

## Generating a Correlation Matrix
```{r apdx2, ref.label="cormat"}
```

## Fitting Models
```{r apdx3, ref.label="fit"}
```

## Presenting Results in Tables
```{r apdx4, ref.label="table"}
```