---
title: "Lab 12"
author: "Haohan Chen"
date: "April 6, 2018"
output: html_document
---

```{r}
#-----------------------------
# load/ install required packages
#-----------------------------
pkgs <- c("lme4", "dplyr", "sandwich", "lmtest", "plm")

usePackage <- function(p) 
{
  for (pkg in p){
    if (!is.element(pkg, installed.packages()[,1]))
      install.packages(pkg, dep = TRUE)
    require(pkg, character.only = TRUE)
  }
}
usePackage(pkgs)
```

# Preparation and Descriptive

Basic model specification
\begin{gather*}
  lnhr_{it} = \beta lnwg_{it} + \epsilon_{it}
\end{gather*}

```{r}
# Load package and data
rm(list=ls())
data(LaborSupply)
data <- LaborSupply
rm(LaborSupply)

# Transform data
pdata <- pdata.frame(data, index = c("id", "year"))

# Variation of hour worked
summary(pdata$lnhr)

# Variation in wage
summary(pdata$lnwg)

# Purpose: Check the across-time variation is large
```

# Pooled model

\begin{gather*}
  y_{it} = \alpha + x_{it} \beta + \epsilon_{it}
\end{gather*}

```{r}
# The pooling model. The simpliest model
m1 <- plm(lnhr ~ lnwg, data = pdata, effect = "individual", model = "pooling")
summary(m1)

# The same as m1.
m1_2 <- lm(lnhr ~ lnwg, data = data)
summary(m1_2)
```

# Between model estimator
\begin{gather*}
  \bar{y}_i = \alpha + \bar{x}_i \beta + \bar{\epsilon}_i
\end{gather*}
```{r}
data %>% group_by(id) %>% summarise_all(funs(mean)) %>% 
  lm(lnhr ~ lnwg, data = .) %>% summary
```

# Heterogeneity via individual specific effects
\begin{gather*}
  y_{it} = \alpha_i + x_{it} \beta + \epsilon_{it}
\end{gather*}

We estiamte different variatns of it

- Fixed effecte
- Random Effect (adding assumption)
\begin{gather*}
  \alpha_i \sim N(0, \sigma_\alpha^2)\\
  \epsilon_{it} \sim N(0, \sigma_\epsilon^2)
\end{gather*}

## Within: Fixed Effect model

\begin{gather*}
  y_{it} = \alpha_i + x_{it} \beta + \epsilon_{it} \\
  \bar{y}_i = \alpha_i + \bar{x}_i \beta + \bar{\epsilon}_{i} \\
  y_{it} - \bar{y}_i = (x_{it} - \bar{x}_i) \beta + (\epsilon_{it} - \bar{\epsilon}_{i}) \\
\end{gather*}

```{r}
# The fixed effect model.
m2 <- plm(lnhr ~ lnwg, data = pdata, effect = "individual", model = "within")
summary(m2)
```

## Within: Fixed Effect with Individual dummies 
```{r}
m2b <- lm(lnhr ~ -1 + lnwg + factor(id), data = data)
summary(m2b)
```

## Random Effect (linear model)
```{r}
m3 <- lmer(lnhr ~ lnwg + (1 | id), data = data)
summary(m3)
```

## Random Effect (Generalized Least Square)
```{r}
m3b <- plm(lnhr ~ lnwg, data = pdata, effect = "individual", model = "random")
summary(m3b)
ercomp(m3b)
```

## First Differences model
```{r}
m4 <- plm(lnhr ~ lnwg, data = pdata, effect = "individual", model="fd")
summary(m4)
```

# Twoway

## Fixed Effect
\begin{gather*}
  y_{it} = \alpha_i + \gamma_t + x_{it} \beta + \epsilon_{it}
\end{gather*}

```{r}
m6 <- plm(lnhr ~ lnwg, data = pdata, effect = "twoway", model = "within")
summary(m6)

# Fixed effect
summary(fixef(m6, effect = "time", type = "level"))
# Deviation from mean
summary(fixef(m6, effect = "time", type = "dmean"))

# Fit it in an naive way
m6b <- lm(lnhr ~ lnwg + factor(year) + factor(id), data = data)
summary(m6b)
```

## Random Effect (GLS)
```{r}
m7 <- plm(lnhr ~ lnwg, data = pdata, effect = "twoway", model = "random")
summary(m7)
```

### Random Effect (Linear model)
```{r}
m7b <- lmer(lnhr ~ lnwg + (1 | id) + (1 | year), data = data)
summary(m7b)
```

# Tests

## Individual specific effects

```{r}
# Recall our pooled model
summary(m1)
plmtest(m1, effect = "individual", type = "bp")
```

`chisq = 2490.1, df = 1, p-value < 2.2e-16`

Reject Null hypothesis that there is no individual specific effect.

## Compare models
```{r}
testw1 <- plm(lnhr ~ lnwg, data = pdata, effect = "individual", model = "within")
pFtest(testw1, m1)
```

`F = 5.8326, df1 = 531, df2 = 4787, p-value < 2.2e-16`

Reject null hypothesis that time specific effect has 0 coefficients. That is, we need to use the within model -- the more complicated one.

# Hausman Test

Choose between fixed effect and random effect model for panel data.
```{r}
fe <- plm(lnhr ~ lnwg, data = pdata, effect = "individual", model = "within")
re <- plm(lnhr ~ lnwg, data = pdata, effect = "individual", model = "random")

phtest(fe, re) # or phtest(re, fe). doesn't matter which one put first
```

`chisq = 13.726, df = 1, p-value = 0.0002115`
Reject the random effect model. "Large test statistics implies rejecting the random effect model"

# Robust Standard Error

```{r}
coeftest(m2)
# Get a more "conservative" estimate of standard error.
# coeftest is from package lmtest. vcovHC is from package sandwich
coeftest(m2, vcov = vcovHC(m2, method="arellano"))
```

# Random Coefficient
\begin{gather*}
  y_{it} = z_{it} \beta_i + \epsilon_{it} \\
  \beta_i = \beta + \alpha_i
\end{gather*}

## FLGS
```{r}
vc <- pvcm(lnhr ~ lnwg, data = pdata, effect = "individual", model = "random")
summary(vc)
```

## Linear Model
```{r}
vcml <- lmer(lnhr ~ lnwg + (1 + lnwg | id), data = data)
summary(vcml)

# This is what it looks like
# a column of intercept and a column of varying slope
# Same as what you see in the previous lab
head(ranef(vcml)$id)
```
